---
title: "Indego Bikeshare Data Exploration"
output: 
  html_document:
    code_folding: show
    theme: flatly
    highlight: breezedark
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r set_dir, include=FALSE}
knitr::opts_knit$set(root.dir = "../data")
```

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(janitor)
library(highcharter)
library(plotly) #https://plotly-r.com/
library(heatmaply)
library(here)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 10)
```

### Initial Look at the Data

```{r}

trips <- read_csv("daily_trips_and_weather.csv") %>%
  rename(trips = n) %>%
  mutate(month = month(date, label = T, abbr = T),
         day_nm = wday(date, label = T, abbr = T)) %>%
  clean_names() 

skimr::skim(trips)

```

The bike share data comes from [Indego](https://www.rideindego.com/about/data/) and according to the documentation its already somewhat cleaned

 > * Staff servicing and test trips are removed.
   * Trips below 1 minute are removed. 
   * A “Virtual Station” listed in the checkout and return kiosks, is used by staff to check in or check out a bike remotely for a special event or in a situation
   in which a bike could not otherwise be checked in or out to a station.
   * Trip lengths are capped at 24 hours.
   * Some short round trips or long trips may be the result of system or user error, but have been kept in the dataset for completeness.

The weather data comes from *DarkSky*, and was obtained using the `darksky` [package.](https://github.com/hrbrmstr/darksky)

Missing data in the weather data set is an issue - `ozone` is 100% missing and `precip_accumulation` is almost entirely missing. Lets remove them as well as any timestamps since they probably won't be useful in the model.

```{r}
trips <- trips %>%
  select(-ends_with("time"), -c(row_num, ozone, precip_accumulation))

skimr::skim(trips)
```

`precip_type` seems to be `NA` when there is no precipitation, so we can recode that to `none`. The handfull of remaining missing values we can omit from the data

```{r}
trips <- trips %>%
  mutate(precip_type = ifelse(is.na(precip_type), "none", precip_type)) %>%
  na.omit() 

skimr::skim(trips)
```


### Fomatting Helper Function
We can put this all together and turn into a `get_trips_data` function. This ensures the data is consistent each time we load it, and can serve as a formatting template for the new weather data we will use in the model when we predict.

This function will live in a `helper.R` file to be used throughout the analysis.

```{r}

get_trips_data <- function(data){
  
  trips <- data %>%
  rename(trips = n) %>%
  mutate(month = month(date, label = T, abbr = T),
         day_nm = wday(date, label = T, abbr = T)) %>%
  clean_names() %>%
  select(-ends_with(c("time", "min", "low")), starts_with("temperature"), -c(row_num, ozone, precip_accumulation, wind_bearing, wind_gust, moon_phase)) %>%
  mutate(precip_type = ifelse(is.na(precip_type), "none", precip_type)) %>%
  na.omit()

  return(trips)
  
}

```

### Data Exploration {.tabset .tabset-fade .tabset-pills}

#### Target
Plotting the daily number of trips over time shows seasonality as well as 1 potential outlier - this could be a data issue.

```{r}

ggplot(trips, aes(x = trips)) + geom_histogram()

ggplot(trips, aes(x = date, y = trips, color = month, group = 1)) + 
  geom_line()

```

It turns out this is when the pope was in Philadelphia, so it is a legitimate data point and we can leave it in!

[pope visits philly](https://www.phillyvoice.com/philadelphias-papal-Sunday/?page=7)

```{r}
trips %>%
  filter(trips == max(trips)) %>%
  pull(date)
```

#### Categorical Predictors
```{r}
categorical_predictors <- trips %>% 
  select_if(., .predicate = is.character) %>%
  names() 

categorical_predictors <- c("trips", categorical_predictors, "day_nm", "month")

get_boxplot <- function(predictor){
  
  df <- trips[names(trips) %in% categorical_predictors]
  
  df <- df %>% 
    pivot_longer(cols = 2:ncol(.)) %>%
    filter(name == {{predictor}}) 
  
  title <- df$name
  
  df %>% 
    ggplot(., aes(x = forcats::fct_reorder(factor(value), trips, .fun = median), y = trips)) +
    geom_boxplot() +
    coord_flip() + 
    labs(x = title, y = "trips", title =  paste('Trips by', title))
}

# check relationship of categorical vars with target
map(categorical_predictors, get_boxplot)
```

#### Numeric Predictors

Based on these scatter plots, it makes sense to remove `moon_phase`,  `wind_bearing` and `wind_gust`. Below the scatterplots is a correlation matrix of predictors. I am also going to remove the `*_min` and `*_low` variables as well since they are highly correlated with their `*_max` and `*_high` counterparts.

Both of these changes are implemented in the `get_trips_data()` function above.

```{r}
# check relationship of numeric vars with target

numeric_predictors <- trips %>% 
  select_if(., .predicate = is.numeric) %>%
  names() 

numeric_predictors <- c("trips", numeric_predictors)

get_scatterplot <- function(predictor){
  
  df <- trips[names(trips) %in% numeric_predictors]
  
  df <- df %>% 
    pivot_longer(cols = 2:ncol(.)) %>%
    filter(name == {{predictor}}) 
  
  title <- df$name
  
  df %>%
    ggplot(., aes(x = value, y = trips)) + 
    geom_point() + 
    geom_smooth() + 
    labs(x = title, y = "trips", title =  paste('Trips vs', title))
}

map(numeric_predictors, get_scatterplot)
```


```{r}
cor_mat <- trips %>% 
  select_if(., is.numeric) %>%
  select(-trips) %>%
  cor()

heatmaply_cor(cor_mat, 
              symm = TRUE, 
              cexRow = .0001, 
              cexCol = .0001, 
              branches_lwd = .1
) 

```

